---
output:
  pdf_document: default
  html_document: default
---

---
title: "Solving Linear Modelds by Brute-Force Part(I)"
author: Abdurrahim Ali.
---
  
# Operations Research
# Laboratory Session 4: Solving linear models with R (part II)
**by Josu Ceberio and Ana Zelaia**

In the previous lab session, the aim was to implement the functions to calculate all the feasible basic solutions of any given linear model. To that end, the definition of the system of linear equations was sufficient, and we did not care about the objective function to accomplish our tasks.

Conversely, in this lab-session, it is our aim to implement the needed functions to solve any given linear model optimally (when it is possible). When the model is feasible, the optimal solution among the basic feasible solutions will be returned. If the model has multiple optimal solutions, then, the function will return all of them. As in the previous lab session, the unbounded problems are out of the scope of the practice.

\begin{eqnarray*}
\max\ \ z= \ \ {\bf c}^{T}{\bf x} \\
\mbox{subject to}\hspace{0,5cm}\\
{\bf A}{\bf x} = {\bf b} \\
{\bf x} \geq {\bf 0}
\end{eqnarray*}

## The `solveProblem` functions

Firstly, given a linear model written in the maximization standard form, the function will verify that the problem is feasible (at least one solution exists for the model). If it is so, then, it will implement the routines that calculate the set of all basic solutions, and using the objective function $z$, return the optimal solution. Specifically, the function must return a list that contains: (1) a vector (if the problem has only one solution) or a matrix (if it has multiple solutions), and (2) the objective value $z^*$ of the optimal solution.

The function will print a message in the standard output providing information related to the set of feasible solutions (their number). If the problem is not feasible, it should print a zero. The `message` function can be very helpful to do this task.

Provide two implementations of the functions: using `for` loops,  and `apply` functions.


To check the correctness of the functions implemented, let us consider the following linear model:
$$
\begin{array}{r}
\max\ z=3x_{1}+4x_{2}+5x_3+6x_4 \\
     \mbox{subject to}\hspace{3cm}\\
      2x_{1}+x_{2}+x_3+8x_4 = 6    \\
      x_{1}+x_{2}+2x_3+x_{4}=4 \\
      x_{1},x_{2},x_{3},x_{4}\geq 0  \\
      \end{array}
$$

```{r, eval=FALSE}
A <- matrix(c(2, 1, 1, 8, 1, 1, 2, 1), nrow=2, byrow=TRUE)
b <- c(6, 4)
c <- c(3, 4, 5, 6)
```

The optimal solution:  $x^*=(0.0000000, 3.7142857, 0.0000000, 0.2857143)$, $z^*=16.57143$


**Exercise 1.** Implementation of `solveProblem` using `for` loops. To ease your task, employ the function `basic.feasible.solutions_for` implemented in the previous laboratory session. Return the result in a list.

```{r, eval=FALSE}
#' solveProblem_for
#' Function takes matrix A, vector b and vector c as input and calculates for each basic feasible solution the z value using a for loop and stores in an auxiliar variable the greatest value calculated for later returning it alongside the vector or vectors that produce it.
#' @param A Matrix holding the vectors to be combined and solved alongside vector b.
#' @param b Vector holding the independent elements of a system.
#' @param c Vector holding the objective function elements.
#'
#' @return List which contains the vector or vectors that produce the optimal solution of the system alongside the calculated value of the optimal solution.
solveProblem_for <- function(A, b, c){
  # Implement here (15 lines >)
  
m <- basic.feasible.solutions_for(A,b)
  optSol <- -Inf
  if(length(m)>0){
    for(i in 1:length(m)){
      tSol <- t(unlist(m[i])) %*% c
      if(tSol>optSol){
        optSol <- tSol
        retSol <- list("solutions" = list(), "optimal.val" = list())
        retSol$solutions <- m[i]
        retSol$optimal.val <- c(optSol)
      }else if(tSol==optSol){
        retSol$solutions <- append(retSol$solutions,m[i])
      }
    }
    return(retSol)
  }else{
    message("Problem has no feasible solutions!")
    message("0")
  }
  
}
solveProblem_for(A,b,c)
# The problem has only one optimal solution.
# $solutions
# $solutions[[1]]
# [1] 0.0000000 3.7142857 0.0000000 0.2857143
# 
# 
# $optimal.val
# [1] 16.57143
```
#' Optimal solution or solutions obtained solving the system alongside the z optimal value.


**Exercise 2.**  Implementation of `solveProblem` using `apply` fuction.  To ease your task, employ the function `basic.feasible.solutions_apply` implemented in the previous laboratory session. Return the result in a matrix.

```{r, eval=FALSE}
#' solveProblem_apply
#' Function takes matrix A, vector b and vector c as input and calculates for each basic feasible solution the z value using the apply() r-function and stores in an auxiliar variable the greatest value calculated for later returning it alongside the vector or vectors that produce it in a list.
#' @param A Matrix holding the vectors to be combined and solved alongside vector b.
#' @param b Vector holding the independent elements of a system.
#' @param c Vector holding the objective function elements.
#'
#' @return List which contains the vector or vectors that produce the optimal solution of the system alongside the calculated value of the optimal solution.

solveProblem_apply <- function(A, b, c){
  # Implement here (15 lines >)
  m <- basic.feasible.solutions_apply(A,b)
  result <- list("solutions"=list(), "optimal_val" = c())
  if(length(m)>0){
    retList <- apply(m, 2, optVal <- function(vec) {
       return(t(unlist(vec)) %*% c)
    })
    
    result$optimal_val <- max(retList)
    
    result$solutions <- apply(m,2, solSelect <- function(vec){
      if((t(unlist(vec)) %*% c)==result$optimal_val){
        return(vec)
      }
    })
    result$solutions<-result$solutions[!sapply(result$solutions,is.null)] #Deleting null values...
    return(result)
  }
}

solveProblem_apply(A,b,c)
# The problem has only one optimal solution.
# $solutions
# [1] 0.0000000 3.7142857 0.0000000 0.2857143
# 
# $optimal.val
# [1] 16.57143
```
#' Optimal solution or solutions obtained solving the system alongside the z optimal value.

#  Linear Models

Use the implemented functions to solve the following problems, and check the correctness of the solutions.

**Problem 1.**  

\vspace{-0.5cm}

\begin{eqnarray*}
\max\ z=-2x_{1}-4x_{2}-3x_3 \\
     \mbox{subject to}\hspace{2.5cm}\\
      2x_{1}+x_{2}+2x_3\geq 8\\
      4x_{1}+2x_{2}+2x_3\geq 10\\
      6x_{1}+x_{2}+4x_3\geq 12\\
      x_{1},x_{2},x_3\geq 0\\
\end{eqnarray*}

\vspace{-0.5cm}

```{r, eval=FALSE}

# Implement here. 
A <- matrix(c(2, 1, 2,-1,0,0, 4, 2, 2,0,-1,0, 6, 1,4,0,0,-1), nrow=3, byrow=TRUE)
b <- c(8,10,12)
c <- c(-2, -4, -3,0,0,0)
solveProblem_apply(A,b,c)
solveProblem_for(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()
```

There is a unique optimal basic feasible solution for the problem. 

$(x^*_1, x^*_2, x^*_3, x^*_4, x^*_5, x^*_6)=(4, 0, 0, 0, 6, 12)$, $z^*=-8$ 


**Problem 2.**  
  
  \vspace{-0.5cm}

\begin{eqnarray*}
\min\ z=2x_{1}+x_{2}+3x_3+2x_4\\
\mbox{subject to}\hspace{3cm}\\
2x_{1}+2x_{2}+2x_3+2x_4\geq 22\\
4x_{1}+4x_{2}+x_3+4x_4\leq 20\\
2x_{1}+8x_{2}+2x_3+x_4\geq 15 \\
x_{1},x_{2},x_{3},x_4\geq 0\\
\end{eqnarray*}

\vspace{-0.5cm}

```{r, eval=FALSE}
A <- matrix(c(2,2,2,2,-1,0,0,4,4,1,4,0,1,0,2,8,2,1,0,0,-1), nrow=3, byrow=TRUE)
b <- c(22,20,15)
c <- c(-2,-1,-3,-2,0,0,0)
sol <- solveProblem_apply(A,b,c)
sol$optimal_val <- sol$optimal_val *-1 #because the standarization
sol
sol<-solveProblem_for(A,b,c)
sol$optimal.val <- sol$optimal.val *-1 #because the standarization
sol
# Implement here.
#' Testing both implemented functions solveProblem_apply() and solveProblem_for()
```

There is a unique optimal basic feasible solution for the problem

$(x^*_1, x^*_2, x^*_3, x^*_4, x^*_5, x^*_6)=(0, 3, 8, 0, 0, 0, 25)$, $z^*=27$ 
  
**Problem 3.**  
  
\vspace{-0.5cm}

\begin{eqnarray*}
\max\ z=x_{1}+2x_{2}\\
\mbox{subject to}\hspace{1.5cm}\\
x_{1}+2x_{2}\leq 5\\
x_{1}+x_{2}\geq 2\\
x_{1}-x_{2}\leq 4 \\
x_{1},x_{2}\geq 0\\
\end{eqnarray*}

There are 2 optimal basic feasible solutions for the problem.

$(x^*_1, x^*_2, x^*_3, x^*_4, x^*_5)=(4.3,\ 0.33,\ 0,\ 2.66,\ 0)$, $z^*= 5$
  
$(x^*_1, x^*_2, x^*_3, x^*_4, x^*_5)=(0.0,\  2.5,\  0.0,\  0.5,\  6.5)$, $z^*= 5$
  
```{r, eval=FALSE}
A <- matrix(c(1,2,1,0,0,1,1,0,-1,0,1,-1,0,0,1), nrow=3, byrow=TRUE)
b <- c(5,2,4)
c <- c(1,2,0,0,0)
# Implement here.
solveProblem_for(A,b,c)

solveProblem_apply(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()
```


**Problem 4.**  
  
\vspace{-0.5cm}

\begin{eqnarray*}
\max\ z=3x_{1}+4x_{2}\\
\mbox{subject to}\hspace{1.5cm}\\
x_{1}-2x_{2}\leq 4\\
x_{1}+x_{2}\geq 6\\
2x_{1}+3x_{2}\leq 2  \\
x_{1},x_{2}\geq 0\\
\end{eqnarray*}

There is no feasible solution for the problem

```{r, eval=FALSE}
A <- matrix(c(1,-2,1,0,0,1,1,0,-1,0,2,3,0,0,1), nrow=3, byrow=TRUE)
b <- c(4,6,2)
c <- c(3,4,0,0,0)
solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)
# Implement here.

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()
```

\pagebreak

# Solving the linear models in the list of Simplex exercises

Solve the problems in the Simplex list of exercises using the functions `solveProblem_for` and `solveProblem_apply`, and verify the correctness of the obtained solutions.

**1.6.1** There is a unique optimal basic feasible solution for the problem.

$x^*= (3.6 \ 0.8\  0.0\  0.0)$, $z^*= 2.8$

```{r, eval=FALSE}

A <- matrix(c(1,-2,1,0,4,-3,0,1), nrow=2, byrow=TRUE)
b <- c(2,12)
c <- c(1,-1,0,0)
# Implement here.
solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)
#' Testing both implemented functions solveProblem_apply() and solveProblem_for()

```


**1.6.2** There are three basic feasible solutions that correspond to the same extreme point ($x^*_1=6, x^*_2=0$).  

$x^*= (6\  0\  0\ 18\  0)$
$x^*= (6\  0\  0\ 18\  0)$
$x^*= (6\  0\  0\ 18\  0)$, $z^*= 6$

```{r, eval=FALSE}
A <- matrix(c(1,6,-1,0,0,2,-3,0,-1,0,1,2,0,0,1), nrow=3, byrow=TRUE)
b <- c(6,-6,6)
c <- c(1,1,0,0,0)
# Implement here.
solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)
#' Testing both implemented functions solveProblem_apply() and solveProblem_for()


```

**1.6.3** There is a unique optimal basic feasible solution for the problem

$x^*= (3.33\  0.33\ 10.00\  0.00\  0.00)$, $z^*= 12$

```{r, eval=FALSE}
A <- matrix(c(-2,2,1,0,0,2,-2,0,1,0,-1,4,0,0,-1), nrow=3, byrow=TRUE)
b <- c(4,6,-2)
c <- c(4,-4,0,0,0)
# Implement here.

solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()

```

**1.6.4** There are 2 optimal basic feasible solutions for the problem.

$x^*= (4.3\ 0.33\ 0\ 2.66\ 0)$, $z^*= 5$
  
$x^*= (0.0\  2.5\  0.0\  0.5\  6.5)$, $z^*= 5$
  
```{r, eval=FALSE}
A <- matrix(c(1,2,1,0,0,1,1,0,-1,0,1,-1,0,0,1), nrow=3, byrow=TRUE)
b <- c(5,2,4)
c <- c(1,2,0,0,0)
# Implement here.

solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()
```

**1.6.5** There are 2 optimal basic feasible solutions for the problem.

$x^*= (1\ 2\ 3\ 0\ 0)$, $z^*= 6$
  
$x^*= (2.5\ 0.5\ 0.0\ 0.0\ 1.5)$, $z^*= 6$
  
```{r, eval=FALSE}
A <- matrix(c(1,-1,1,0,0,2,2,0,1,0,1,2,0,0,1), nrow=3, byrow=TRUE)
b <- c(2,6,5)
c <- c(2,2,0,0,0)
# Implement here.
solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()

```

**1.6.6** The problem is unbounded. Do not do.

**1.6.7** The problem is unbounded. Do not do.

**1.6.8** There is no feasible solution for the problem

```{r, eval=FALSE}
A <- matrix(c(1,-2,1,0,0,1,1,0,-1,0,2,3,0,0,1), nrow=3, byrow=TRUE)
b <- c(4,6,2)
c <- c(3,4,0,0,0)
# Implement here.
solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()

```

**1.7.1** There is a unique optimal basic feasible solution for the problem

$x^*= (0\  8\  0\ 12\  0)$, $z^*= 16$
  
```{r, eval=FALSE}
A <- matrix(c(1,-1,1,1,0,2,1,4,0,1), nrow=2, byrow=TRUE)
b <- c(4,8)
c <- c(3,2,1,0,0)
# Implement here.

solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()

```

**1.7.2** There is a unique optimal basic feasible solution for the problem

$x^*= (0\ 6.666\ 0.6666\ 8.6666\ 0)$, $z^*= 8$. Sign change...

$x^*= (0\ 6.666\ 0.6666\ 8.6666\ 0)$, $z^*= -8$.

```{r, eval=FALSE}
A <- matrix(c(1,1,-1,0,0,0,1,2,-1,0,-1,0,-1,1,2,0,0,1), nrow=3, byrow=TRUE)
b <- c(6,4,8)
c <- c(-5,1,2,0,0,0)
# Implement here.
sol <- solveProblem_apply(A,b,c)
sol$optimal_val <- sol$optimal_val *-1 #because the standarization
sol
sol<-solveProblem_for(A,b,c)
sol$optimal.val <- sol$optimal.val *-1 #because the standarization
sol

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()
```


**1.7.3** There is a unique optimal basic feasible solution for the problem

$x^*= (1.625\ 0\ 2.0625\ 0\ 15.1875\ 0)$, $z^*= 5.75$.
  
```{r, eval=FALSE}
A <- matrix(c(3,-1,2,1,0,0,-2,4,1,0,1,0,-4,4,8,0,0,1), nrow=3, byrow=TRUE)
b <- c(9,14,10)
c <- c(1,-3,2,0,0,0)
# Implement here.
solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()
```

**1.7.4** There are 2 optimal basic feasible solutions for the problem

$x^*= (0\ 2\ 0\ 2\ 0\ 0)$, $x^*= (0\ 3\ 0\ 0\ 2\ 0)$, $z^*= -24$. 

Sign change... $z^*= 24$
  
```{r, eval=FALSE}
A <- matrix(c(2,4,2,1,-1,0,-4,4,-1,2,0,-1), nrow=2, byrow=TRUE)
b <- c(10,12)
c <- c(-10,-8,-6,-4,0,0)
# Implement here.
sol <- solveProblem_apply(A,b,c)
sol$optimal_val <- sol$optimal_val *-1 #because the standarization
sol
sol<-solveProblem_for(A,b,c)
sol$optimal.val <- sol$optimal.val *-1 #because the standarization
sol

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()

```

**1.7.5** There is no feasible solution for the problem.

```{r, eval=FALSE}
A <- matrix(c(2,1,1,2,1,0,0,8,4,-2,-1,0,-1,0,4,7,2,1,0,0,1), nrow=3, byrow=TRUE)
b <- c(2,10,4)
c <- c(-9,-5,-4,-1,0,0,0)
# Implement here.

solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()


```

**1.7.6** There are 2 optimal basic feasible solutions for the problem

$x^*= (4.0\ 0.0\ 0.0\ 1.0\ 0.0\ 0.0\ 0.0)$, 

$x^*= (0\  0\  0\ 21\  8\  0\  0)$, $z^*= 66$. 

Sign change... $z^*= -66$
  
```{r, eval=FALSE}

A <- matrix(c(3,1,3,-3,9,1,0,2,8,4,2,-4,0,1), nrow=2, byrow=TRUE)
b <- c(9,10)
c <- c(16,-2,-1,2,3,0,0)

# Implement here.
sol <- solveProblem_apply(A,b,c)
sol$optimal_val <- sol$optimal_val *-1 #because the standarization
sol
sol<-solveProblem_for(A,b,c)
sol$optimal.val <- sol$optimal.val *-1 #because the standarization
sol

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()

```

**1.7.7** Unbounded problem. Do not do this exercise.

**1.7.8** There is no feasible solution for the problem.

```{r, eval=FALSE}
A <- matrix(c(1,2,2,1,1,1,0,2,1,3,2,2,0,-1), nrow=2, byrow=TRUE)
b <- c(2,12)
c <- c(-3,-1,2,2,-1,0,0)

# Implement here.

solveProblem_apply(A,b,c)

solveProblem_for(A,b,c)

#' Testing both implemented functions solveProblem_apply() and solveProblem_for()

```
